HealpixNet: Efficient spherical Convolutional Neural Network with HEALPix sampling for cosmological applications

### 2018-08-12 scnn.tex
nati pass 1st part

% cohen2017convolutional (ICML workshop) and cohen2018sphericalcnn (ICLR best paper) are redundant
A first direction to generalize CNNs to spherical data was to apply a standard 2D CNN to a grid discretisation of the sphere~\citep{boomsma2017spherical,su2017sphericalconv,coors2018spherenet}. While this approach uses the well-developed 2D convolution and hierarchical pooling, it is restricted to those specific grid samplings which are highly irregular on the sphere and hence inherently deform the domain where the data is supported. 
A second direction attempted by~\citet{cohen2018sphericalcnn} was to leverage the spherical Fourier transform and to perform the convolution associated to the SO(3) rotation group in the spectral domain, thanks to the convolution theorem.
Contrarily to the first direction, the resulting convolution is rotational equivariant, meaning that a rotation of the input implies the same rotation of the output.
This is an important property as translation invariance is probably one of the main reason why classic CNN were successful in the first place.
Nevertheless, this approach is computationally expensive even if a fast spherical Fourier transforms is used. 
%(They exist for particular samplings.)
Finally, the use of a graph to model the discretised sphere was considered by~\citet{khasanova2017graphomni} as well, towards the goal of designing network that were invariant to rotations.
While their architecture accommodates these invariances on the sphere and remains fairly efficient, they did not take advantage of a hierarchical pixelisation and use a grid discretisation of the sphere. 
Moreover, the statistical layer, which introduces the invariance, needs to be hand-crafted to capture statistics of interest.

@@

A CNN is composed of the following main building blocks~\citep{lecun1998cnn}:
(i) a convolution,
(ii) a non-linearity,
(iii) a down-sampling operation,
(iv) a pooling operation, and
(v), optionally,  normalization.\footnote{\citet{ioffe2015batchnorm} have shown that batch normalization helps training. We verified this experimentally in our setting as well.}
Our architecture is depicted in~\figref{architecture} and discussed in greater details in \secref{architecture}. As operations (ii) and (v) are point-wise, they do not depend on the data domain. The pooling operation is simply a permutation invariant aggregation function which does not need to be adapted either. The convolution and down-sampling operations however need to be generalized from Euclidean domains to the sphere.

@@

HEALPix is an acronym for Hierarchical Equal Area isoLatitude Pixelization of a sphere.
This pixelisation produces a hierarchical subdivision of a spherical surface where each pixel covers the same surface area as every other pixel.
A hierarchy is desired for data locality in computer memory.
Equal area is advantageous because white noise generated by the signal receiver gets integrated exactly into white noise in the pixel space.
Isolatitude is essential for the implementation of a fast spherical transform.
HEALPix is the sole pixelisation scheme which satisfies those three properties.

The lowest possible resolution is given by the base partitioning of the surface into $N_{pix} = 12$ equal sized pixels (right-most sphere in \figref{pooling}).
As each pixel is subdivided in four, the second coarser resolution is $N_{pix} = N_{side}^2 \cdot 12 = 2^2 \cdot 12 = 48$ pixels (middle sphere in \figref{pooling}), the third is $N_{pix} = N_{side}^2 \cdot 12 = 4^2 \cdot 12 = 192$ pixels, etc., where $N_{side} = 1, 2, 4, 8, \ldots$ is the grid resolution parameter.
High-resolutions HEALPix maps easily reach millions of pixels.
\figref{example_maps} shows two examples of HEALPix maps: the Cosmic Microwave Background \citep{planck2015overview}, galaxies found in Sloan Digital Sky Survey Data Release 14 \citep{abolfathi2017sdssDR14}, and an example simulated mass map on the footprint of Dark Energy Survey Data Release 1 \citep{des2018dr1}.

@@

Our graph is constructed as an approximation of the continuous 2D manifold that is the surface of the sphere.
Indeed, \citet{belkin2007convergence} showed that the graph Laplacian converges to the Laplace-Beltrami when the number of pixels goes to infinity.
That construction can accommodate any sampling, regular or not.

>>>>

% cohen2017convolutional (ICML workshop) and cohen2018sphericalcnn (ICLR best paper) are redundant
A first direction to generalize CNNs to spherical data was to apply a standard 2D CNN to a grid discretisation of the sphere~\citep{boomsma2017spherical,su2017sphericalconv,coors2018spherenet}. While this approach uses the well-developed 2D convolution and hierarchical pooling, it is restricted to those specific grid samplings which are highly irregular on the sphere and hence inherently deform the domain where the data is supported. 
A second direction attempted by~\citet{cohen2018sphericalcnn} was to leverage the spherical Fourier transform and to perform the convolution associated to the SO(3) rotation group in the spectral domain, thanks to the convolution theorem.
Contrarily to the first direction, the resulting convolution is rotational equivariant, meaning that a rotation of the input implies the same rotation of the output.
This is an important property as translation invariance is probably one of the main reasons why classic CNNs are successful in the first place.
Nevertheless, this approach is computationally expensive even if a fast spherical Fourier transforms is used. 
%(They exist for particular samplings.)
Finally, the use of a graph to model the discretised sphere was considered by~\citet{khasanova2017graphomni} as well, towards the goal of designing a  network that is invariant to rotations.
While their architecture accommodates these invariances on the sphere and remains fairly efficient, they did not take advantage of a hierarchical pixelisation and use as pixelisation a grid discretisation of the sphere. 
Moreover, the statistical layer, which introduces the invariance, needs to be hand-crafted to capture statistics of interest.

@@

A CNN is composed of the following main building blocks~\citep{lecun1998cnn}:
(i) a convolution,
(ii) a non-linearity,
(iii) a down-sampling operation,
(iv) a pooling operation, and
(v), optionally,  normalization.\footnote{\citet{ioffe2015batchnorm} have shown that batch normalization helps training. We verified this experimentally in our setting as well.}
Our architecture is depicted in~\figref{architecture} and discussed in greater details in \secref{architecture}. As operations (ii) and (v) are point-wise, they do not depend on the data domain. The pooling operation is simply a permutation invariant aggregation function which does not need to be adapted either. The convolution and down-sampling operations, however, need to be generalized from Euclidean domains to the sphere.

@@

HEALPix is an acronym for Hierarchical Equal Area isoLatitude Pixelization of a sphere.
This pixelisation produces a hierarchical subdivision of a spherical surface where each pixel covers the same surface area as every other pixel.
A hierarchy is desired for the data locality in the computer memory.
Equal area is advantageous because white noise generated by the signal receiver gets integrated exactly into white noise in the pixel space.
Isolatitude is essential for the implementation of a fast spherical transform.
HEALPix is the sole pixelisation scheme which satisfies those three properties.

The lowest possible resolution is given by the base partitioning of the surface into $N_{pix} = 12$ equal-sized pixels (right-most sphere in \figref{pooling}).
As each pixel is subdivided in four, the second coarser resolution is $N_{pix} = N_{side}^2 \cdot 12 = 2^2 \cdot 12 = 48$ pixels (middle sphere in \figref{pooling}), the third is $N_{pix} = N_{side}^2 \cdot 12 = 4^2 \cdot 12 = 192$ pixels, etc., where $N_{side} = 1, 2, 4, 8, \ldots$ is the grid resolution parameter.
High-resolutions HEALPix maps easily reach millions of pixels.
\figref{example_maps} shows two examples of HEALPix maps: the Cosmic Microwave Background \citep{planck2015overview}, galaxies found in Sloan Digital Sky Survey Data Release 14 \citep{abolfathi2017sdssDR14}, and an example simulated mass map on the footprint of Dark Energy Survey Data Release 1 \citep{des2018dr1}.

@@

Our graph is constructed as an approximation of the continuous 2D manifold that is the surface of the sphere.
Indeed, \citet{belkin2007convergence} showed that the graph Laplacian converges to the Laplace-Beltrami when the number of pixels goes to infinity.
While our construction will not exactly respect the setting defined by \citet{belkin2007convergence}, we observe empirically strong evidence of convergence.
